Specific taxa differentiate proximal and distal microbiota in the unprepped

### 2017-06-27 submission/manuscript.Rmd
added in some Pat edits

stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
leftandrighttyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
lvsr <- rbind(stooltyc, leftandrighttyc)

@@

#####RandomForest###########################################################################################
#build randomForest model for each location comparison using randomize_loc function 
rf_left <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
rf_right <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
rf_bowel <- randomize_loc(rel_meta, "LB", "RB") #OOB 25.64%
rf_lumen <- randomize_loc(rel_meta, "LS", "RS") #OOB 69.23%

@@

# create RF model with AUCRF outputs top OTUs
aucrf_data_left_bs <- auc_loc(rel_meta, "LB", "LS")
aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
aucrf_data_right_bs <- auc_loc(rel_meta, "RB", "RS")
aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")

@@

#10fold CV for L lumen vs L mucosa
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp_left_bs <- c()
cv10f_all_pred_left_bs <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data_left_bs),nrow(aucrf_data_left_bs),replace=F)
  cv10f_probs <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_left_bs <- aucrf_data_left_bs[sampling[-(i:(i+3))],]
    test_left_bs <- aucrf_data_left_bs[sampling[i:(i+3)],]
    set.seed(seed)
    temp_model_left_bs <- AUCRF(location~., data=train_left_bs, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_left_bs$RFopt, test_left_bs, type='prob')[,2]
  }
  cv10f_roc_left_bs <- roc(aucrf_data_left_bs$location~cv10f_probs)
  cv10f_all_pred_left_bs <- c(cv10f_all_pred_left_bs, cv10f_probs)
  cv10f_all_resp_left_bs <- c(cv10f_all_resp_left_bs, aucrf_data_left_bs$location)
  cv10f_aucs[j] <- cv10f_roc_left_bs$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_left_bs <- roc(cv10f_all_resp_left_bs~cv10f_all_pred_left_bs)

#10fold CV for R lumen vs R mucosa
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp_right_bs <- c()
cv10f_all_pred_right_bs <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data_right_bs),nrow(aucrf_data_right_bs),replace=F)
  cv10f_probs <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_right_bs <- aucrf_data_right_bs[sampling[-(i:(i+3))],]
    test_right_bs <- aucrf_data_right_bs[sampling[i:(i+3)],]
    set.seed(seed)
    temp_model_right_bs <- AUCRF(location~., data=train_right_bs, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_right_bs$RFopt, test_right_bs, type='prob')[,2]
  }
  cv10f_roc_right_bs <- roc(aucrf_data_right_bs$location~cv10f_probs)
  cv10f_all_pred_right_bs <- c(cv10f_all_pred_right_bs, cv10f_probs)
  cv10f_all_resp_right_bs <- c(cv10f_all_resp_right_bs, aucrf_data_right_bs$location)
  cv10f_aucs[j] <- cv10f_roc_right_bs$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_right_bs <- roc(cv10f_all_resp_right_bs~cv10f_all_pred_right_bs)

@@

#10 fold cross validation for L vs R lumen
iters <- 100
cv10f_aucs_lum <- c()
cv10f_all_resp_lum <- c()
cv10f_all_pred_lum <- c()
for(j in 1:iters){
  set.seed(j)
  sampling_lum <- sample(1:nrow(aucrf_data_LRlumen),nrow(aucrf_data_LRlumen),replace=F)
  cv10f_probs_lum <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_lum <- aucrf_data_LRlumen[sampling_lum[-(i:(i+3))],]
    test_lum <- aucrf_data_LRlumen[sampling_lum[i:(i+3)],]
    set.seed(seed)
    temp_model_lum <- AUCRF(location~., data=train_lum, pdel=0.99, ntree=500)
    cv10f_probs_lum[sampling_lum[i:(i+3)]] <- predict(temp_model_lum$RFopt, test_lum, type='prob')[,2]
  }
  cv10f_roc_lum <- roc(aucrf_data_LRlumen$location~cv10f_probs_lum)
  cv10f_all_pred_lum <- c(cv10f_all_pred_lum, cv10f_probs_lum)
  cv10f_all_resp_lum <- c(cv10f_all_resp_lum, aucrf_data_LRlumen$location)
  cv10f_aucs_lum[j] <- cv10f_roc_lum$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_lum <- roc(cv10f_all_resp_lum~cv10f_all_pred_lum)
```
```{r important_otus, echo=FALSE}
#just LB vs LS 
left_otu_feat <- colnames(aucrf_data_left_bs[2:6])
otu_taxa_left <- get_tax(1, left_otu_feat, tax_file)
#Abundance stripchart or most predictive otus
ls_abunds <- shared_meta[shared_meta$location=='LS', left_otu_feat]/10000 + 1e-4
lb_abunds <- shared_meta[shared_meta$location=='LB', left_otu_feat]/10000 + 1e-4

right_otu_feat <- colnames(aucrf_data_right_bs[2:6])
otu_taxa_right <- get_tax(1, right_otu_feat, tax_file)
#Abundance stripchart or most predictive otus
rs_abunds <- shared_meta[shared_meta$location=='RS', right_otu_feat]/10000 + 1e-4
rb_abunds <- shared_meta[shared_meta$location=='RB', right_otu_feat]/10000 + 1e-4

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures. Tumors of the proximal (right) and distal (left) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. _Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The left mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the left lumen, again reflecting the anatomical structure of the colon. By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the left side of the colon are infiltrating lesions that present with painful symptoms. In contrast, 47% of CRCs are caused by right-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010). Due to the absence of symptoms, right-sided tumors have a significantly higher mortality rate (@Benedix2010). The left and right sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016).  Because these studies were performed in mice or examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined. Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct left and right CRC tumors.

@@

We combined these approaches to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon. Our design used unprepared colonoscopy techniques to sample each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address noise created by a diverse set of human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Proximal and distal lumenal and mucosal samples were collected from 20 healthy humans that had not undergone bowel preparation(Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. The relative abundance of each phyla for each site is shown in Figure 2A. This data combines the samples from all participants in the study. Each site is primarily dominated by _Firmicutes_ and _Bacteriodetes_, consistent with known variability in human microbiome research (cite). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). That is, while some individuals had a more diverse right mucosa, some had a more diverse left mucosa. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (left or right) or sites (lumen or mucosa), we calculated theta YC distances from OTU abundances and compared distances for all individuals. Again, across all patient samples we observed a range of theta YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the right lumen and mucosa, the left versus right lumen, the left versus right mucosa and the left lumen and mucosa, we found that the right lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

Next, we calculated thetaYC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller thetaYc distance between the left lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the left mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were overall much more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013). Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results of the similarity analysis suggest that the contents of the left lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models built on OTU abundances. We built the first model to classify the lumen versus mucosal samples for the right and left sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the right and ((Xopt)) for the left. The models performed well when classifying these samples (`r cv10f_roc_right_bs$auc` and `r cv10f_roc_left_bs$auc`, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the right lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the left lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the left and right microbiome. The model performed best when distinguishing the right versus left mucosa (Figure 5A, AUC = 0.912) compared to the right versus left lumen (AUC = 0.755). The model was able to explain ((X%)) of the variance. OTUs that were differentially abundant between the left and right mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). The model was less-effective at classifying the right and left lumen (AUC = 0.755). However it did identify differentially abundant OTUs such as three members of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also drive sidedness- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated wtih the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_. It was the only sequence to align with a _F. nucleatum_ strain- the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located on the right and left mucosa and ranged up to 1% of the sample (Figure 6A). OTU00152 is _Porphyromonas_ and the most frequent sequence in that OTU aligns to _Porphyromonas asacharolytica_, another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the left mucosa, and in fact was one of the OTUs the classification model identified as separating left and right sides (Figure 6B). Samples that were positive for _P. asacharolytica_ ranged in abundances from 0.01% - 16%. Thus, cancer-associated OTUs could be found in our sample set of 20 healthy individuals.

### Discussion
Here we identified bacterial taxa that are specific to the lumen and mucosa of the right and left human colon from samples collected during unprepared colonoscopy. We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects. Pairwise comparisons of each of the sites revealed that the right mucosa and lumen were most similar to each other. Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the left lumen is most similar to stool at exit. Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon. Finally, we were able to detect some bacterial OTUs associated with colorectal cancer in our healthy patient cohort. Using unprepped colonoscopy and machine learning, we have identified bacterial phyla specific to the healthy proximal and distal human colon.

>>>>

stooltyc <- subset(tyc, match=='LB_RB' | match== 'LS_RS')
distalandproximaltyc <- subset(tyc, match=='LB_LS' | match== 'RB_RS')
lvsr <- rbind(stooltyc, distalandproximaltyc)

@@

#####RandomForest###########################################################################################
#build randomForest model for each location comparison using randomize_loc function 
rf_distal <- randomize_loc(rel_meta, "LB", "LS") #OOB 10.26%
rf_proximal <- randomize_loc(rel_meta, "RB", "RS") #OOB 53%
rf_bowel <- randomize_loc(rel_meta, "LB", "RB") #OOB 25.64%
rf_lumen <- randomize_loc(rel_meta, "LS", "RS") #OOB 69.23%

@@

# create RF model with AUCRF outputs top OTUs
aucrf_data_distal_bs <- auc_loc(rel_meta, "LB", "LS")
aucrf_data_LRbowel <- auc_loc(rel_meta, "LB", "RB")
aucrf_data_proximal_bs <- auc_loc(rel_meta, "RB", "RS")
aucrf_data_LRlumen <- auc_loc(rel_meta, "LS", "RS")
aucrf_data_allum <- auc_site(rel_meta, "mucosa", "stool")

@@

#10fold CV for L lumen vs L mucosa
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp_distal_bs <- c()
cv10f_all_pred_distal_bs <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data_distal_bs),nrow(aucrf_data_distal_bs),replace=F)
  cv10f_probs <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_distal_bs <- aucrf_data_distal_bs[sampling[-(i:(i+3))],]
    test_distal_bs <- aucrf_data_distal_bs[sampling[i:(i+3)],]
    set.seed(seed)
    temp_model_distal_bs <- AUCRF(location~., data=train_distal_bs, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_distal_bs$RFopt, test_distal_bs, type='prob')[,2]
  }
  cv10f_roc_distal_bs <- roc(aucrf_data_distal_bs$location~cv10f_probs)
  cv10f_all_pred_distal_bs <- c(cv10f_all_pred_distal_bs, cv10f_probs)
  cv10f_all_resp_distal_bs <- c(cv10f_all_resp_distal_bs, aucrf_data_distal_bs$location)
  cv10f_aucs[j] <- cv10f_roc_distal_bs$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_distal_bs <- roc(cv10f_all_resp_distal_bs~cv10f_all_pred_distal_bs)

#10fold CV for R lumen vs R mucosa
iters <- 100
cv10f_aucs <- c()
cv10f_all_resp_proximal_bs <- c()
cv10f_all_pred_proximal_bs <- c()
for(j in 1:iters){
  set.seed(j)
  sampling <- sample(1:nrow(aucrf_data_proximal_bs),nrow(aucrf_data_proximal_bs),replace=F)
  cv10f_probs <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_proximal_bs <- aucrf_data_proximal_bs[sampling[-(i:(i+3))],]
    test_proximal_bs <- aucrf_data_proximal_bs[sampling[i:(i+3)],]
    set.seed(seed)
    temp_model_proximal_bs <- AUCRF(location~., data=train_proximal_bs, pdel=0.99, ntree=500)
    cv10f_probs[sampling[i:(i+3)]] <- predict(temp_model_proximal_bs$RFopt, test_proximal_bs, type='prob')[,2]
  }
  cv10f_roc_proximal_bs <- roc(aucrf_data_proximal_bs$location~cv10f_probs)
  cv10f_all_pred_proximal_bs <- c(cv10f_all_pred_proximal_bs, cv10f_probs)
  cv10f_all_resp_proximal_bs <- c(cv10f_all_resp_proximal_bs, aucrf_data_proximal_bs$location)
  cv10f_aucs[j] <- cv10f_roc_proximal_bs$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_proximal_bs <- roc(cv10f_all_resp_proximal_bs~cv10f_all_pred_proximal_bs)

@@

#10 fold cross validation for L vs R lumen
iters <- 100
cv10f_aucs_lum <- c()
cv10f_all_resp_lum <- c()
cv10f_all_pred_lum <- c()
for(j in 1:iters){
  set.seed(j)
  sampling_lum <- sample(1:nrow(aucrf_data_LRlumen),nrow(aucrf_data_LRlumen),replace=F)
  cv10f_probs_lum <- rep(NA,39)
  for(i in seq(1,36,4)){
    train_lum <- aucrf_data_LRlumen[sampling_lum[-(i:(i+3))],]
    test_lum <- aucrf_data_LRlumen[sampling_lum[i:(i+3)],]
    set.seed(seed)
    temp_model_lum <- AUCRF(location~., data=train_lum, pdel=0.99, ntree=500)
    cv10f_probs_lum[sampling_lum[i:(i+3)]] <- predict(temp_model_lum$RFopt, test_lum, type='prob')[,2]
  }
  cv10f_roc_lum <- roc(aucrf_data_LRlumen$location~cv10f_probs_lum)
  cv10f_all_pred_lum <- c(cv10f_all_pred_lum, cv10f_probs_lum)
  cv10f_all_resp_lum <- c(cv10f_all_resp_lum, aucrf_data_LRlumen$location)
  cv10f_aucs_lum[j] <- cv10f_roc_lum$auc #stores aucs for all iterations, can use to calc IQR
}
cv10f_roc_lum <- roc(cv10f_all_resp_lum~cv10f_all_pred_lum)
```
```{r important_otus, echo=FALSE}
#just LB vs LS 
distal_otu_feat <- colnames(aucrf_data_distal_bs[2:6])
otu_taxa_distal <- get_tax(1, distal_otu_feat, tax_file)
#Abundance stripchart or most predictive otus
ls_abunds <- shared_meta[shared_meta$location=='LS', distal_otu_feat]/10000 + 1e-4
lb_abunds <- shared_meta[shared_meta$location=='LB', distal_otu_feat]/10000 + 1e-4

proximal_otu_feat <- colnames(aucrf_data_proximal_bs[2:6])
otu_taxa_proximal <- get_tax(1, proximal_otu_feat, tax_file)
#Abundance stripchart or most predictive otus
rs_abunds <- shared_meta[shared_meta$location=='RS', proximal_otu_feat]/10000 + 1e-4
rb_abunds <- shared_meta[shared_meta$location=='RB', proximal_otu_feat]/10000 + 1e-4

@@

Colorectal cancer (CRC) remains a leading cause of death worldwide despite improved preventative and therapeutic measures. Tumors of the proximal (proximal) and distal (distal) colon are morphologically and genetically distinct. Previous work from our group found that microbial dysbiosis is associated with the development of CRC tumors in studies of both mice and humans. Analysis of the fecal microbiota from healthy and CRC patients further revealed different microbial signatures associated with disease. We extended our observations of the fecal microbiome to include analysis of the proximal and distal human colon. We used a two-colonoscope approach on subjects that had not undergone standard bowel preparation procedure. This technique allowed us to characterize the native proximal and distal luminal and mucosal microbiome without prior chemical disruption. 16S rRNA gene sequencing was performed on proximal and distal mucosal and luminal biopsies and home-collected stool for 20 healthy individuals. Diversity analysis revealed that each site contained a diverse community, and that a patient’s samples were more similar to each other than to that of other individuals. Since we could not differentiate sites along the colon based on community structure or community membership alone, we employed the Random Forest machine-learning algorithm to identify key species that distinguish biogeographical sites. Random Forest classification models were built using taxa abundance and sample location and revealed distinct populations that were found in each location. _Peptoniphilus, Anaerococcus, Enterobacteraceae, Pseudomonas_ and _Actinomyces_ were most likely to be found in mucosal samples versus luminal samples (AUC = 0.925). The classification model performed well (AUC = 0.912) when classifying mucosal samples into proximal or distal sides, but separating luminal samples from each side proved more challenging (AUC = 0.755). The distal mucosa was found to have high populations of _Finegoldia, Murdochiella_ and _Porphyromonas_. Proximal and distal luminal samples were comprised of many of the same taxa, likely reflecting the fact that stool moves along the colon from the proximal to distal end. Finally, comparison of all samples to fecal samples taken at exit uncovered that the feces were most similar to samples taken from the distal lumen, again reflecting the anatomical structure of the colon. By sampling the unprepped human colon, our results have identified distinct bacterial populations native to the proximal and distal sides. Further investigation of these bacteria may elucidate if and how these groups contribute to differential oncogenesis processes on the respective sides of the colon.

@@

Colorectal cancer (CRC) is the second-leading cause of cancer-related deaths in the United States. CRC tumors vary in structure, size, morbidity and symptomology depending upon their geographical location in the colon. CRC has recently been described to fall into four molecular subtypes (@Guinney2015). Of these subtypes, tumors that arise on the distal side of the colon are infiltrating lesions that present with painful symptoms. In contrast, 47% of CRCs are caused by proximal-sided colon tumors that are bulky and project into the lumen, often remaining asymptomatic until advanced carcinogenesis (@Benedix2010). Due to the absence of symptoms, proximal-sided tumors have a significantly higher mortality rate (@Benedix2010). The distal and proximal sides of the colon differ in the amount of inflammation present and the genomic instability of precancerous cells, respectively, as well as oxygen, pH and the presence of antimicrobial peptides (@Yamauchi2012, @Albenberg2014, @Donaldson2015). Microenvironments differ not only longitutinally along the colon, but latitudinally from the epithelium to mucosa to intestinal lumen, offering several sites for different microbial communities to flourish. Given this varied physiology of the proximal-distal axis of the colon, symbiotic microbes and their metabolites likely vary between sites.

Several recent findings have shown that development and progression of CRC can be attributed to specific molecular events as a result of interactions between the gut microbiota and human host (@Yamauchi2012). Our group and others have found that the stool microbiome of patients with CRC is distinct from that of healthy people (@Zackular2014). Further studies manipulating the gut microbiome using antibiotics or other chemoagents in mice has shown that dysbiosis preceded and accelerated the development of CRC tumors (@Zackular2013). Comparison of the specific bacteria present on CRC tumors with those found on nearby healthy tissue has also identified specific bacterial species that are tumor-associated (@Dejea2014). These species include the oral pathogens _Fusobacterium nucleatum_ and _Porphyromonas asacharolytica_. Interestingly, these periodontal pathogens have been highly predictive of whether a patient had CRC tumors or not in our classification studies (@Baxter2016).  Because these studies were performed in mice or examined only shed human stool, they were unable to analyze paried samples from the proximal and distal sides of the colon. Similarly, comparisons of on- or off-tumor bacteria rarely have matched tissue from the other side of the colon from the same patient, limiting what conclusions can be drawn about the colonic microbiome overall, let alone at that specific site. Due to these limitations, the contribution of the gut microbiota to these subtypes is largely undefined. Characterizing these communities could provide needed insight into CRC etiology, including how the disruption of the healthy community could promote the initiation or proliferation of the distinct distal and proximal CRC tumors.

@@

Here we aimed to address the limitations of previous studies and effectively characterize the microbiome in the lumen and mucosa of the proximal and distal healthy human colon. Our design used unprepared colonoscopy techniques to sample the natural community of each location of the gut without prior disruption of the native bacteria in 20 healthy volunteers. To address the inherent inter-individual variation in human microbiomes, we used a machine-learning classification algorithm trained on curated 16S rRNA sequencing reads to identify microbes specific to each location. We found that our classification models were able to separate mucosal and lumenal samples as well as differentiate between sides of the colon based on populations of specific microbes. By identifying the specific microbes we are poised to ask if and how the presence or disruption of the microbes at each site contribute to the development of the specific tumor subtypes of CRC in the proximal and distal human colon.

@@

#### Microbial membership and diversity of the proximal and distal colon 
Lumenal and mucosal samples were collected from the proximal and distal colon of 20 healthy humans that had not undergone bowel preparation (Figure 1). Participants also collected stool at home one week prior to the procedure. To characterize the bacterial communities present at these sites, 16S rRNA gene sequencing was performed on extracted DNA from each sample. Each site was primarily dominated by _Firmicutes_ and _Bacteriodetes_ (Figure 2A), consistent with known variability in human microbiome research (cite). Likewise, samples had varying levels of diversity at each site, irrespective of the individual (Figure 2B). For example, the proximal mucosa was more diverse than the distal for some individuals while the opposite was true for others. Therefore we could not identify a clear pattern of changes in microbial diversity along the gut axis.

To compare similarity between sides (proximal or distal) or sites (lumen or mucosa), we calculated ${\theta}$YC distances from OTU abundances and compared these distances for all individuals. Again, across all patient samples we observed a range of ${\theta}$YC distances when comparing sample locations (Figure 3A) and again those ranges did not follow a clear pattern on an individual basis. However, when comparing median distances between the proximal lumen and mucosa, the proximal versus distal lumen, the proximal versus distal mucosa, and the distal lumen and mucosa, we found that the proximal lumen and mucosa were most similar to each other than the other samples (P < 0.005, Wilcoxon, BH adjustment).

#### Stool at exit most resembles lumenal samples from the distal colon 
Next, we calculated ${\theta}$YC distances to examine how each sample compared to the home-collected exit stool. Amidst variability between patients, we did identify significantly smaller ${\theta}$YC distance between the distal lumenal sample and the exit stool (Figure 3B, P < 0.05, Wilcoxon, BH adjustment). Furthermore, there was an even larger difference in the comparisons of the distal mucosa to the exit stool, indicating that the mucosa is different from the stool as compared to lumen (P < 0.0005, Wilcoxon, BH adjustment). To determine what factors may be driving the differences seen among the samples, we compared thetaYC distances between samples from all subjects (interpersonal) versus samples from within one subject (intrapersonal). We found that samples from one individual were far more similar to each other than to other study subjects (Figure 3C), consistent with previous human microbiome studies that have sampled multiple sites of the human colon (@Eckberg2005, @deCarcer2010, @Zhang2013). Thus interpersonal variation between subjects drives the differences between samples more than sample site or location. Overall, the results comparing the structure of the communities suggest that the contents of the distal lumen are most representative of stool at exit, and the microbes remaining on the mucosa are much different.

#### Random Forest classification models identify important OTUs on each side
To identify OTUs that were distinct at each biogeographical site, we constructed several Random Forest models trained using OTU abundances. We built the first model to classify the lumen versus mucosal samples for the proximal and distal sides, independently (Figure 4A). The constructed model used ((Xopt)) features for the proximal and ((Xopt)) for the distal. The models performed well when classifying these samples (`r cv10f_roc_proximal_bs$auc` and `r cv10f_roc_distal_bs$auc`, respectively). The OTUs that were most predictive of each site are identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs from the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and two OTUs from the _Enterobacteraceae_ genera were differentially abundant (Figure 4B). The model classifying the distal lumen and mucosa identified OTUs from _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and two OTUs from the _Anaerococcus_ genera that could distinguish lumen from mucosal samples (Figure 4C). These results indicate that there are fine differences between the different sites of the colon, and that these can be traced down to specific OTUs on each side.

Next, we built a model to differentiate the proximal and distal lumenal samples. The model performed best when distinguishing the proximal versus distal mucosa (Figure 5A, AUC = 0.912) compared to the proximal versus distal lumen (AUC = 0.755). These models were able to explain ((X%)) of the variance, respectively. OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5B). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5C). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (XX% over full length). This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. (@Castellarin2011). The _Fusobacterium_ positive samples were located in x% of the the proximal and X% of the distal mucosa and represented as much as 1% of any sample (Figure 6A). OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (X% over full length), another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure 6B). Among the samples that were positive for _P. asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%. Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.

### Discussion
Here we identified bacterial taxa that are specific to the lumen and mucosa of the proximal and distal human colon from samples collected during unprepared colonoscopy. We found that all locations contained a range of phyla and a range of diversity, but that there was a wide variability between subjects. Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other. Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen is most similar to stool at exit. Random Forest algorithms built on OTU abundances from each sample identified microbes that are particular to each location of the colon. Finally, we were able to detect some bacterial OTUs associated with colorectal cancer in our healthy patient cohort. Using unprepped colonoscopy and machine learning, we have identified bacterial phyla specific to the healthy proximal and distal human colon.
